---
title: "Filter ATAC-Seq Data"
author: "Eleanor Alspaugh , Max Alcer, Jakub Liu"
date: "2024-06-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(Matrix)
```
# Load Expression Matrix and annotations
DO NOT ATEMT TO RUN WITH <64GB OF RAM
```{r}
expression_matrix <- readRDS("../data/unprocessed/ATAC_counts.rds") # not included in git
```

```{r}
annotations <- read_csv("../data/unprocessed/OMIX928-24.csv", show_col_types = FALSE)
```

```{r}
annotations <- annotations %>% select(Barcode, Celltype) %>%
  filter(grepl("Erythroblast", Celltype)|
           grepl("Cycling_Blood_Progenitor", Celltype)|
           grepl("Macrophage", Celltype)|
           grepl("Megakaryocyte_Progenitor", Celltype)|
           grepl("Myeloid_Cell", Celltype)|
           grepl("Blood_Progenitor", Celltype)) %>%
  mutate(Barcode = sub(paste0(".*\\", "#"), "", Barcode))
  #mutate(Celltype = replace(Celltype, grepl("Erythroblast", Celltype), "Erythroblast")) %>%
  #mutate(Celltype = replace(Celltype, grepl("Cycling_Blood_Progenitor", Celltype), "Cycling_Blood_Progenitor")) %>% 
  #mutate(Celltype = replace(Celltype, grepl("Macrophage", Celltype), "Macrophage")) %>% 
  #mutate(Celltype = replace(Celltype, grepl("Megakaryocyte_Progenitor", Celltype), "Megakaryocyte_Progenitor")) %>%
  #mutate(Celltype = replace(Celltype, grepl("Myeloid_Cell", Celltype), "Myeloid_Cell")) %>%
  #mutate(Celltype = replace(Celltype, grepl("Blood_Progenitor", Celltype), "Blood_Progenitor"))
```

```{r}
write_csv(annotations, "../data/pre_processed/ATAC_annotations.csv")
```


```{r}
expression_matrix <- expression_matrix[, annotations$Barcode]
```

```{r}
row_max <- apply(expression_matrix, 1, max)
indx_genes <- which(row_max > max(row_max)/50)
expression_matrix <- expression_matrix[indx_genes,]
```

```{r}
saveRDS(expression_matrix, file = "../data/pre_processed/ATAC-Seq-filtered.rds", compress = TRUE)
```





