---
title: "Filter ATAC-Seq Data"
author: "Eleanor Alspaugh , Max Alcer, Jakub Liu"
date: "2024-06-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(Matrix)
```
# Load Expression Matrix and annotations
DO NOT ATEMT TO RUN WITH <64GB OF RAM
```{r}
expression_matrix <- readRDS("../data/unprocessed/ATAC_counts.rds") # not included in git
```

```{r}
annotations <- read_csv("../data/unprocessed/OMIX928-24.csv", show_col_types = FALSE)
```


# Only keep the cells releated to Haematopoiesis and add a column that 
# represents the sub-trajectory from the MOCA dataset 
```{r}
annotations <- annotations %>% select(Barcode, Celltype) %>%
  filter(grepl("Erythroblast", Celltype)|
           grepl("Macrophage", Celltype)|
           grepl("Megakaryocyte_Progenitor", Celltype)|
           grepl("Myeloid_Cell", Celltype)|
           grepl("Blood_Progenitor", Celltype)) %>%
  mutate(Barcode = sub(paste0(".*\\", "#"), "", Barcode)) %>%
  mutate(Trajectory = case_when(
    grepl("Erythroblast", Celltype) & grepl("G01", Celltype) ~ "primitive erythroid trajectory",
    grepl("Erythroblast", Celltype) & !grepl("G01", Celltype) ~ "Definitive erythroid trajectory",
    grepl("Blood_Progenitor", Celltype) ~ "primitive erythroid trajectory",
    grepl("Macrophage", Celltype) ~ "White blood cell trajectory",
    grepl("Myeloid_Cell", Celltype) ~ "White blood cell trajectory",
    grepl("Megakaryocyte_Progenitor", Celltype) ~ "Megakaryocyte trajectory",
  ))
  
```

```{r}
write_csv(annotations, "../data/pre_processed/ATAC_annotations.csv")
```

# Only keep cells in the expression matrix that are releated to Haematopoiesis
# and filter out genes where the highest expression is lower than 2% of the 
# highest expression value
```{r}
expression_matrix <- expression_matrix[, annotations$Barcode]
```

```{r}
row_max <- apply(expression_matrix, 1, max)
indx_genes <- which(row_max > max(row_max)/50)
expression_matrix <- expression_matrix[indx_genes,]
```

```{r}
saveRDS(expression_matrix, file = "../data/pre_processed/ATAC-Seq-filtered.rds", compress = TRUE)
```





