---
title: "Filter ATAC-Seq Data"
author: "Eleanor Alspaugh , Max Alcer, Jakub Liu"
date: "2024-06-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(Matrix)
```


# Load Expression Matrix and annotations
DO NOT ATEMT TO RUN WITH <64GB OF RAM
```{r}
expression_matrix <- readRDS("../data/unprocessed/gene_count_cleaned.RDS") # not included in git
```

```{r}
annotations <- read_csv("../data/unprocessed/cell_annotate.csv", show_col_types = FALSE)
```


```{r}
annotations <- annotations %>% 
  filter(Sub_trajectory_name %in% c("Definitive erythroid trajectory",
                                    "White blood cell trajectory",
                                    "primitive erythroid trajectory",
                                    "Megakaryocyte trajectory"),
         Main_trajectory == "Haematopoiesis trajectory",
         Total_mRNAs >= 400,
         doublet_cluster == FALSE,
         detected_doublet == FALSE)%>%
  select(sample, Main_cell_type, Main_trajectory, Sub_trajectory_name)
  
  
```

```{r}
write_csv(annotations, "../data/pre_processed/RNA_annotations.csv")
```



```{r}
expression_matrix <- expression_matrix[,annotations$sample]
```

```{r}
row_max <- apply(expression_matrix, 1, max)
indx_genes <- which(row_max > 10)
expression_matrix <- expression_matrix[indx_genes,]
```

```{r}
saveRDS(expression_matrix, file = "../data/pre_processed/RNA-Seq-filtered.rds")
```





